### 2019-05-11 papers/haskell19/3-representation.tex
Last minute changes

For this purpose, we will define a set of different simple ``open'' types, each
one in charge of representing a single random construction from our
\emph{target} data type, i.e., the actual data type we want to randomly
generate.
%
These types can be later
%
\begin{inparaenum}[(i)]
\item combined in several ways depending on the desired shape of our test data
  (applying the familiar \`a la Carte technique);
\item randomly generated (see Section \ref{sec:generators}); and finally,
\item transformed back to the corresponding values of the target data type
  automatically.
\end{inparaenum}
%
This lifted representation of constructions can be automatically derived from
our source code in compile-time, relieving programmers of the burden of manually
implementing the required machinery.

@@

%
% Closing
%
\subsubsection{Ensambling Altogether}

Using the constructions presented so far, we are now able to represent values of
our target data using a higher level of abstraction.
%
Instead of constructing values solely by combining data constructors, this
mechanism lets us also consider complex constructions like abstract interface
functions and functions' patterns, treating them uniformingly and in a fine
grained fashion.

>>>>

For this purpose, we will define a set of different simple ``open'' types, each
one in charge of representing a single random construction from our
\emph{target} data type, i.e., the actual data type we want to randomly
generate.
%
These types can be later
%
\begin{inparaenum}[(i)]
\item combined in several ways depending on the desired shape of our test data
  (applying the familiar \`a la Carte technique);
\item randomly generated (see Section \ref{sec:generators}); and finally,
\item transformed back to the corresponding values of the target data type
  automatically.
\end{inparaenum}
%
This lifted representation of constructions can be automatically derived from
our source code at compile time, relieving programmers of the burden of manually
implementing the required machinery.

@@

%
% Closing
%
\subsubsection{Assembling It All Together}

Using the constructions presented so far, we are now able to represent values of
our target data using a higher level of abstraction.
%
Instead of constructing values solely by combining data constructors, this
mechanism lets us also consider complex constructions like abstract interface
functions and functions' patterns, treating them uniformly and in a fine grained
fashion.

### 2019-05-11 papers/haskell19/4-generators.tex
Minor changes based on Nachi's feedback
NIT
Last minute changes

Ideally, our aim is to be able to obtain random generators with a behavior
similar to the one presented for |Html| in Section \ref{sec:intro}.
%
If we take a closer look at its definition, we can observe three factors
happening simultaneously:
%
\vspace{2pt}
\begin{CompactItemize}
\item We use QuickCheck's generation size to limit the depth of the generated
  values, reducing it by one on each recursive call of the local auxiliary
  function |gen|.
\item We differentiate between \emph{terminal} and \emph{non-terminal
    constructors}, picking only among terminal ones when we have reached the
  maximum depth (case |gen 0|).
\item We generate different constructions in a (possibly) different frequency.
\end{CompactItemize}
%
For the rest of this section, we will focus on modeling these aspects in our
modular framework, in such a way that does not compromise the compositionality
obtained so far.

@@

instance BArbitrary1 Con_Join where
  liftBGen bgen d = Mk_Join <$> bgen (d - 1) <*> bgen (d - 1)
\end{code} %$

Note how each instance is defined to be parametric of the maximum depth (using
the input integer |d|) and the random generator used for the recursive sub-terms
(using the input generator |bgen|).
%
Every other non-recursive sub-term can be generated using a normal |Arbitrary|
instance---we use this to generate random |String|s in the previous definitions.

@@

Recalling the random generator for |Html| presented in Section \ref{sec:intro},
we can observe that the last generation level (see |gen 0|) is constrained to
generate values only from the subset of terminal constructions.
%
In order to model this behavior, we will first defined a data type |Term| to
``tag'' every terminal construction explicitly:

@@

Moreover, if it happens that both |FreqOf' f| and |FreqOf' g| compute to zero
simultaneously, then this will produce a runtime error triggered by the function
|frequency|, as it does not have anything with a positive frequency to generate.
%
This kind of exceptions will arise, for example, if we forget to tag at least a
one construction from our final representation as terminal.
%
Fortunately, such runtime exceptions can be catched at compile-time.
%
We can define a type constraint |HasTerminals| that ensures we are trying to
generate values using a representation with a strictly positive terminal
generation frequency---thus containing at least a single terminal construction:

>>>>

Ideally, our aim is to be able to obtain random generators with a behavior
similar to the one presented for |Html| in Section \ref{sec:intro}.
%
If we take a closer look at its definition, there we can observe three factors
happening simultaneously:
%
\vspace{2pt}
\begin{CompactItemize}
\item We use QuickCheck's generation size to limit the depth of the generated
  values, reducing it by one on each recursive call of the local auxiliary
  function |gen|.
\item We differentiated between \emph{terminal} and \emph{non-terminal
    constructors}, picking only among terminal ones when we have reached the
  maximum depth (case |gen 0|).
\item We generate different constructions in a (possibly) different frequency.
\end{CompactItemize}
%
For the rest of this section, we will focus on modeling these aspects in our
modular framework, in such a way that does not compromise the compositionality
obtained so far.

@@

instance BArbitrary1 Con_Join where
  liftBGen bgen d = Mk_Join <$> bgen (d - 1) <*> bgen (d - 1)
\end{code} %$
%
Note how each instance is defined to be parametric of the maximum depth (using
the input integer |d|) and the random generator used for the recursive sub-terms
(using the input generator |bgen|).
%
Every other non-recursive sub-term can be generated using a normal |Arbitrary|
instance---we use this to generate random |String|s in the previous definitions.

@@

Recalling the random generator for |Html| presented in Section \ref{sec:intro},
we can observe that the last generation level (see |gen 0|) is constrained to
generate values only from the subset of terminal constructions.
%
In order to model this behavior, we will first define a data type |Term| to
``tag'' every terminal construction explicitly:

@@

Moreover, if it happens that both |FreqOf' f| and |FreqOf' g| compute to zero
simultaneously, then this will produce a runtime error triggered by the function
|frequency|, as it does not have anything with a positive frequency to generate.
%
This kind of exceptions will arise, for example, if we forget to tag at least a
one construction from our final representation as terminal.
%
Fortunately, such runtime exceptions can be caught at compile time.
%
We can define a type constraint |HasTerminals| that ensures we are trying to
generate values using a representation with a strictly positive terminal
generation frequency---thus containing at least a single terminal construction:

### 2019-05-11 papers/haskell19/5-typelevel.tex
Minor changes based on Nachi's feedback
NIT
Last minute changes

This naming convention results hard to use, specially if we consider that we do
not know the actual type names until they are synthesized in compile-time, due
to their unique suffixes.
%
Fortunately, it is easy to solve this problem using some type-level machinery.
%
Instead of imposing a naming convention in our derivation tool, we define a set
of open type families to hide each kind of construction behind meaningful names:

@@

\begin{code}
type Html_Spec  =      Con'  "Text"
                oplus  Con   "Join"         otimes 2
                oplus  Fun'  "br"           otimes 4
                oplus  Fun   "<+>"          otimes 9
                oplus  Pat'  "simplify"  1  otimes 3
                oplus  Pat   "simplify"  2  otimes 5
\end{code}

And simply obtain a concrete generator by referring to |Html_Spec| when using
|genRep|:

@@

% The biggest different appears when deriving the required type instances for the
% |Con|, |Fun| and |Pat| synonyms.
%
However, instead of carrying this type parameter in our generation
specifications, we can avoid it by hiding it behind an existential type:

@@

\begin{code}
type instance Con "Leaf" = Some Con_Leaf
type instance Con "Node" = Some Con_Node
\end{code}

And we can therefore write generation specifications for our |Tree| data type
without having to refer to its type parameter anywhere.
%
For instance:

>>>>

This naming convention results hard to use, specially if we consider that we do
not know the actual type names until they are synthesized during compilation,
due to their unique suffixes.
%
Fortunately, it is easy to solve this problem using some type-level machinery.
%
Instead of imposing a naming convention in our derivation tool, we define a set
of open type families to hide each kind of construction behind meaningful names:

@@

\begin{code}
type Html_Spec  =      Con'  "Text"
                oplus  Con   "Join"         otimes 2
                oplus  Fun'  "br"           otimes 4
                oplus  Fun   "<+>"          otimes 9
                oplus  Pat'  "simplify"  1  otimes 3
                oplus  Pat   "simplify"  2  otimes 5
\end{code}
%
And simply obtain a concrete generator by referring to |Html_Spec| when using
|genRep|:

@@

% The biggest different appears when deriving the required type instances for
% the |Con|, |Fun| and |Pat| synonyms.
%
However, instead of carrying this type parameter in our generation
specifications, we can avoid it by hiding it behind an existential type:

@@

\begin{code}
type instance Con "Leaf" = Some Con_Leaf
type instance Con "Node" = Some Con_Node
\end{code}
%
And as a consequence, we can write generation specifications for our |Tree| data
type without having to refer to its type parameter anywhere.
%
For instance:

### 2019-05-11 papers/haskell19/8-conclusions.tex
Last minute changes

We presented a novel approach for automatically deriving flexible random
generators inspired by the seminal work on Data Types \`a la Carte.
%
Furthermore, we incorporate valuable structural information into our generation
process by considering not only data constructors, but also the structural
information statically available in abstract interfaces and functions' pattern
matchings.

In the future, we aim to minimize the runtime overhead of our technique, using
perhaps prefix-free encodings based on generation frequencies, as a way to
reduce the representation bureaucracy as much as possible.
%
Additionally, we are interested on including additional sources of structural
information from the users' codebases, e.g., locally defined patterns or
sequences of functions applications---interesting thoughts to motivate our
future research.

>>>>

We presented a novel approach for automatically deriving flexible random
generators inspired by the seminal work on Data Types \`a la Carte.
%
Furthermore, we incorporate valuable structural information into our generation
process by considering not only data constructors, but also the structural
information statically available in abstract interfaces and functions' pattern
matchings.
%

In the future, we aim to minimize the runtime overhead of our technique, using
perhaps prefix-free encodings based on generation frequencies, as a way to
reduce the representation bureaucracy as much as possible.
%
In addition, using type-level representations for each random construction
leaves the door open for extending our approach with a composable mechanism for
predicting the distributions of random values.
%
This would be possible by overloading each representation type with stochastic
information describing its branching behavior---interesting thoughts to motivate
our future research.

% Additionally, we are interested on including additional sources of structural
% information from the users' codebases, e.g., locally defined patterns or
% sequences of functions applications

### 2019-05-11 papers/haskell19/abstract.tex
Last minute changes

Generating ``good'' random values described by algebraic data types is often
quite intricate.
%
State-of-the-art tools for synthesizing random generators serve the valuable
purpose of helping with this task, while providing different level of invariants
imposed over the generated values.
%
However, none of them are built for composability nor extensibility, a useful
feature when the shape of our random data makes a difference while testing
different properties or sub-systems.
%
% Instead, users are forced to decide wether to manually write their random
% generators or to use rigid automatically derived ones.

In this work, we develop an extensible framework for deriving compositional
generators, which can be easily combined in different ways in order to fit the
developers' demands using a simple type-level description language.
%
Our framework relies on familiar ideas from the \`a la Carte technique for
writing composable interpreters in Haskell.
%
In particular, we adapt this technique with the machinery required in the scope
of random generation.
%
We provide an implementation of our ideas, and evaluate its performance using
real-world examples.
%

>>>>

Generating ``good'' random values described by algebraic data types is often
quite intricate.
%
State-of-the-art tools for synthesizing random generators serve the valuable
purpose of helping with this task, while providing different levels of
invariants imposed over the generated values.
%
However, none of them are built for composability nor extensibility, a useful
feature when the shape of our random data makes a difference while testing
different properties or sub-systems.
%
% Instead, users are forced to decide wether to manually write their random
% generators or to use rigid automatically derived ones.

In this work, we develop an extensible framework for deriving compositional
generators, which can be easily combined in different ways in order to fit
developers' demands using a simple type-level description language.
%
Our framework relies on familiar ideas from the \`a la Carte technique for
writing composable interpreters in Haskell.
%
In particular, we adapt this technique with the machinery required in the scope
of random generation.
%
We provide an implementation of our ideas, and evaluate its performance using
real-world examples.
%
